---
title: "Data preprocessing and creation of the data objects pasillaGenes and pasillaExons"
author: "Alejandro Reyes"
date: "`r doc_date()`"
package: "`r pkg_ver('pasilla')`"
output:
  BiocStyle::html_document:
    toc: true
bibliography: library.bib
vignette: >
  %\VignetteIndexEntry{"Data preprocessing and creation of the data objects pasillaGenes and pasillaExons"}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
abstract: >
  This vignette describes the steps that were followed for the generation 
  of the data objects contained in the package `r Biocpkg("pasilla")`.
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(tidy = FALSE, cache = TRUE, autodep = TRUE)
```

# Downloading the files

We used the RNA-Seq data from the publication by Brooks et al. [@Brooks2010].
The experiment investigated the effect of siRNA knock-down of pasilla, a gene 
that is known to bind to mRNA in the spliceosome, and which is thought to be involved 
in the regulation of splicing.  The data set contains 3 biological replicates of the 
knockdown as well as 4 biological replicates for the untreated control. 
The data files are available from the NCBI Gene Expression Omnibus under the 
[accession GSE18508](http://www.ncbi.nlm.nih.gov/projects/geo/query/acc.cgi?acc=GSE18508). 
The read sequences in FASTQ format were extracted from the NCBI short read archive file 
(.sra files) using the SRA toolkit.

# Read alignment and filtering

The reads in the FASTQ files were aligned using `tophat` version 1.2.0
with default parameters against the reference Drosophila melanogaster
genome.  The following table summarizes the read number and
alignment statistics.
The column `exon.counts` refers to the number of reads that could be uniquely aligned to an exon.

```{r samples}
dataFilesDir = system.file("extdata", package = "pasilla", mustWork=TRUE)
pasillaSampleAnno = read.csv(file.path(dataFilesDir, "pasilla_sample_annotation.csv"))
pasillaSampleAnno
```

The reference genome fasta files were obtained from the 
[Ensembl ftp server](http://www.ensembl.org/info/data/ftp/index.html). 
We ran `bowtie-build` to index the
fasta file. For more information on this procedure see
the [`bowtie` webpage](http://bowtie-bio.sourceforge.net/index.shtml).
The indexed form is required by `bowtie`, and thus `tophat`. 

```{sh wget, eval = FALSE}
wget ftp://ftp.ensembl.org/pub/release-62/fasta/drosophila_melanogaster/ \
dna/Drosophila_melanogaster.BDGP5.25.62.dna_rm.toplevel.fa.gz

gunzip Drosophila_melanogaster.BDGP5.25.62.dna_rm.toplevel.fa.gz
bowtie-build Drosophila_melanogaster.BDGP5.25.62.dna_rm.toplevel.fa \
    d_melanogaster_BDGP5.25.62
```

We generated the alignment BAM file using `tophat`.  For the single-reads data:

```{sh tophat1, eval = FALSE}
tophat bowtie_index <filename>1.fastq,<filename>2.fastq,...,<filename>N.fastq
```
  
For the paired-end data:

```{sh tophat2, eval = FALSE}
tophat -r inner-fragment-size bowtie_index \
    <filename>1_1.fastq,<filename>2_1.fastq,...,<filename>N_1.fastq \
    <filename>1_2.fastq,<filename>2_2.fastq,...,<filename>N_2.fastq
```

The SAM alignment files from which `r Biocpkg("pasilla")` was generated 
[are available at this URL](http://www-huber.embl.de/pub/DEXSeq/analysis/brooksetal/bam). 


# Exon count files

To generate the per-exon read counts, we first needed to define the exonic regions.  
To this end, we downloaded the file `Drosophila_melanogaster.BDGP5.25.62.gtf.gz` from 
[Ensembl](ftp://ftp.ensembl.org/pub/release-62/gtf/drosophila_melanogaster).
The script `dexseq_prepare_annotation.py` contained in the `r Biocpkg("DEXSeq")`
package was used to extract the exons of the transcripts from the file, define new 
non-overlapping exonic regions and reformat it to create the file `Dmel.BDGP5.25.62.DEXSeq.chr.gff`
contained in `pasilla/extdata`. For example, for this file we ran:

```{sh dexseqprep, eval = FALSE}
wget ftp://ftp.ensembl.org/pub/release-62/gtf/ \
drosophila_melanogaster/Drosophila_melanogaster.BDGP5.25.62.gtf.gz

gunzip Drosophila_melanogaster.BDGP5.25.62.gtf.gz
python dexseq_prepare_annotation.py Drosophila_melanogaster.BDGP5.25.62.gtf \
    Dmel.BDGP5.25.62.DEXSeq.chr.gff
```

To count the reads that fell into each non-overlapping exonic part, the script 
`dexseq_count.py`, which is also contained in the `r Biocpkg("DEXSeq")` package, was used. 
It took the alignment results in the form of a SAM file
(sorted by position in the case of a paired end data) and the `gtf` file 
`Dmel.BDGP5.25.62.DEXSeq.chr.gff` and returned one file for each biological replicate 
with the exon counts. For example, for the file treated1.bam, which contained single-end
alignments, we ran:

```{sh dexseqcount1, eval = FALSE}
samtools index treated1.bam
samtools view treated1.bam > treated1.sam
python dexseq_count.py Dmel.BDGP5.25.62.DEXSeq.chr.gff \
    treated1.sam treated1fb.txt
```

For the file `treated2.bam`, which contained paired-end alignments:

```{sh dexseqcount2, eval = FALSE}
samtools index treated2.bam
samtools view treated2.bam > treated2.sam
sort -k1,1 -k2,2n treated2.sam > treated2_sorted.sam
python dexseq_count.py -p yes Dmel.BDGP5.25.62.DEXSeq.chr.gff \
    treated2_sorted.sam treated2fb.txt
```

The output of the two HTSeq python scripts is provided in the 
`r Biocpkg("pasilla")` package:

```{r dirextData}
dir(system.file("extdata", package = "pasilla", mustWork=TRUE), pattern = ".txt$")
```
 
The Python scripts are built upon the
[HTSeq library](http://www-huber.embl.de/users/anders/HTSeq/doc/overview.html).
 
# Creation of the DEXSeqDataSet dxd

To create a DEXSeqDataSet object, we need the count files, the sample annotations,
and the GFF file with the per exon annotation.

```{r ecs}
gffFile = file.path(dataFilesDir, "Dmel.BDGP5.25.62.DEXSeq.chr.gff")
```

With these, we could call the function `DEXSeqDataSet` to construct the object `dxd`.

```{r read, message = FALSE, warning = FALSE}
library("DEXSeq")
dx = DEXSeqDataSetFromHTSeq(
  countfiles = file.path(dataFilesDir, paste(pasillaSampleAnno$file, "txt", sep=".")), 
  sampleData = pasillaSampleAnno,
  design= ~ sample + exon + condition:exon,
  flattenedfile = gffFile)
dx  
```

```{r weird, echo = FALSE}
stopifnot(ncol(dx)==14)
```
<font color="red">FIXME: why does dx have 14 rows and not 7?</font>


We only wanted to work with data from a subset of genes, which was defined in the following file.


```{r dxd}
genesforsubset = readLines(file.path(dataFilesDir, "geneIDsinsubset.txt"))
dxd = dx[geneIDs( dxd ) %in% genesforsubset,]
```

We saved the R objects in the data directory of the package:
```{r save, eval = FALSE}
save(dxd, file = file.path("..", "data", "pasillaDEXSeqDataSet.RData"))
```

# SessionInfo

```{r sessionInfo}
sessionInfo()
```


# References
